Selection of Polynomial Features by Genetic Algorithms

### 2013-07-23 texts/paper_spfga-final.tex
Pre-submission: * renamed and reencoded main .tex file; * archived old text

%
%	FILE ENCODING
%
\usepackage[latin1]{inputenc}
%

@@

@@

\title{Selection of Polynomial Features by Genetic Algorithms}
\author[ue,labmag]{Francisco Coelho}
\ead{fc@di.uevora.pt}

\author[fcul,labmag]{João Pedro Neto}
\ead{jpn@di.fc.ul.pt}

\address[ue]{Dept. Informática, Universidade de Évora, Rua Romão Ramalho 58 Évora 7000-671 Évora}
\address[fcul]{Dept. Informática, Faculdade de Ciências da Universidade de Lisboa, Campo Grande 1749-016 Lisboa}
\address[labmag]{Laboratory of Agent Modelling (LabMAg)}

@@

\end{frontmatter}

@@

One interesting take on genetic algorithms, named \emph{genetic programming} by John Koza \cite{koza1992genetic}, proposed the use of \acp{GA} to search the syntactic structure of complex functions. This syntatic structure search is keen to the central ideas of deep learning \cite{bengio2013representation, bengio2009learning}, a subarea of machine learning actually producing quite promising results (\emph{e.g.} \cite{tarlowstochastic}). It is also related to the work presented in this paper in the sense that, unlike linear models that have a simple structure, $y=\sum_i \beta_i x_i$, nonlinear (in particular polynomial) models pose an additional ``structure" search problem.

The idea of using \acp{GA} to find a polynomial regression is not new \cite{maertens2006genetic, yu2008optimal, wu2009novel} but still generates original research \cite{hofwing2011optimal, cetisli2011polynomial}. In line with that research this work describes a general method to find a polynomial regression of a given dataset. The optimal regression minimises a cost function that accounts for both the \ac{rmse} and a regularization factor to avoid over-fitting.

@@

\item find the fittest set of monomials, \emph{i.e.} the polynomial that minimizes the \ac{rmse} on the same dataset;
\end{enumerate}
%
More precisely, concerning the first problem, let $\mathcal{D}$ be a dataset with $n$ observations of variables $Y, X_1,\ldots,X_m$ and $\mathcal{P} = \left\lbrace q_1,\ldots, q_k\right\rbrace$ a set of $k$ monomial expressions over $X_1,\ldots,X_m$. Define the hypothesis\footnote{The expression $q|_{X=x}$ reads ``\emph{replace all instances of $X$ by $x$ in $q$}''.}
%
\begin{eqnarray*}
h_{\Theta,\mathcal{P}}\left(x_1,\ldots,x_m\right) &=& \sum_{j = 1}^k \theta_j q_j|_{X_i=x_i,\forall 1 \leq i \leq m}
\end{eqnarray*}
%
and let the cost
%
\begin{eqnarray}
J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right) &=& \sqrt{\frac{1}{n}\sum_{i=1}^n \left( y\sample{i} - h_{\Theta,\mathcal{P}}\left( x_1\sample{i},\ldots,x_m\sample{i} \right) \right)^2 }\label{eq:rmse}
\end{eqnarray}
%
be the usual \acf{rmse} function. Now the first problem can be stated as:
%
\begin{quote}
Given a dataset $\mathcal{D}$ and a set of monomials $\mathcal{P}$, find parameters $\Theta$ that minimize $J\left(\Theta;\mathcal{P},\mathcal{D}\right)$.
\end{quote}
%
It turns out that this problem can be solved as a usual linear regression problem by expanding $\mathcal{D}$ with columns that replicate the monomials in $\mathcal{M}$.

The second problem is treated in the \ac{GA} setting: Let $\mathcal{D}$ be a dataset as above and $\mathcal{Q}$ a set of polynomials. For each polynomial $p\in \mathcal{Q}$ let $\mathcal{M}_p$ be the set of monomials in $p$ (without the coefficients) and compute the fitness
%
\begin{eqnarray*}
\phi_p &=& \min_\Theta J\left(\Theta;\mathcal{M}_p,\mathcal{D}\right)
\end{eqnarray*}
%
by solving the first problem. With a fitness of every instance, a \ac{GA} will apply genetic operators (usually mutation and crossover) to evolve the population $\mathcal{P}$ until a reasonable approximation of a local minimum is found. 
%
\begin{algorithm}[tb]
\begin{algorithmic}
\Function{GAPoly}{$D,pop_0,\epsilon$}
	\State $pop \gets pop_0;\: err \gets 1.0+\epsilon$
	\While{$err > \epsilon$}
		\State $pop \gets \Call{IterateGA}{pop}$\Comment{Build next generation}
		\State $pop \gets \Call{Sort}{pop, key = \textsc{Lm.rmse}}$\Comment{Sort by \ac{rmse} of linear regression}
		\State $err \gets \textsc{rmse}\left(\Call{First}{pop}\right)$
	\EndWhile
	\State\Return $\Call{First}{pop}$
\EndFunction
\end{algorithmic}
\caption{\ac{GAPoly} uses linear regression to find monomial coefficients that minimize the \ac{rmse} over a dataset and \acp{GA} to explore the space of polynomials. At exit the \ac{rmse} of the fittest instance is bounded by $\epsilon$.}\label{alg:gapoly}
\end{algorithm} 
%
Notice that the properties of \acp{GA} and linear regression entail that
the composition of \acp{GA} with linear regression, as defined in Algorithm \ref{alg:gapoly}, converges to a polynomial that is a local minimum of the fitness function, encapsulated in the \ac{rmse} function $J_{\textrm{fit}}$.
%

@@

This dataset was used in order to verify if \ac{GAPoly} was able to find the polynomial relation, which the algorithm did (cf. figure~\ref{artificial_dataset1_lambda1.0}).The genetic algorithm run with a population of $N=100$~solutions with 50~iterations for each run.

@@

\item{\textsc{Housing}}: This data set concerns the task of predicting housing values in areas of Boston. There are $m=13$ continuous attributes and the dependent variable is the median value of owner-occupied homes in \$1000's. There are $N=506$ observations.

@@

The proposed method has competitive results comparing with some well-known regression methods. Only Random Forest outperforms \ac{GAPoly} systematically (it also outperforms all the other regression algorithms). One exception -- the artificial dataset that uses a straightforward polynomial relation -- is the Auto-MPG datatest where \ac{GAPoly} has comparable results. This is evidence that applying standard genetic operators for polynomial model searching is a viable tool for regression purposes.

\ac{GAPoly} demands some processing time. On a current quad-core computer, processing the Kinematics dataset (with 8k~observations) takes approximately 5~minutes. The processing time can probably be speeded by one to two orders in magnitude if the process is implemented in a low level programming language like C++. However, speed optimization was not the focus of this article.

@@

The authors are grateful to the \textsc{Fundação para a Ciência e Tecnologia} (FCT) and the  R\&D laboratory \textsc{LabMAg} for the financial support given to this work, under the strategic project \textsc{PEst-­OE/EEI/UI0434/2011}.

The datasets used herein were selected from Luis Torgo's data  repository, www.dcc.fc.up.pt/~ltorgo/Regression/DataSets.html. Most of these datasets originally come from UCI ML repository, http://archive.ics.uci.edu/ml/.

>>>>

%
%	FILE ENCODING
%
\usepackage[utf8]{inputenc}
%

@@

%
%	HYPERREFERENCES
%
\usepackage{hyperref}
%

@@

\title{Selection of Polynomial Features by Genetic Algorithms}

\author[ue,labmag]{Francisco Coelho}
\ead{fc@di.uevora.pt}

\author[fcul,labmag]{João Pedro Neto\corref{jpn}}
\ead{jpn@di.fc.ul.pt}
\cortext[jpn]{Corresponding author}

\address[ue]{Dept. Informática, Universidade de Évora, Rua Romão Ramalho 58, 7000-671 Évora}
\address[fcul]{Dept. Informática, Faculdade de Ciências da Universidade de Lisboa, Campo Grande 1749-016 Lisboa}
\address[labmag]{Laboratory of Agent Modelling (LabMAg)}

@@

\begin{keyword}
polynomial regression \sep genetic algorithm 
\end{keyword}
\end{frontmatter}

@@

One interesting take on genetic algorithms, named \emph{genetic programming} by John Koza \cite{koza1992genetic}, proposed the use of \acp{GA} to search the syntactic structure of complex functions. This syntactic structure search is keen to the central ideas of deep learning \cite{bengio2013representation, bengio2009learning}, a subarea of machine learning actually producing quite promising results (\emph{e.g.} \cite{tarlowstochastic}). It is also related to the work presented in this paper in the sense that, unlike linear models that have a simple structure, $y=\sum_i \beta_i x_i$, nonlinear (in particular polynomial) models pose an additional ``structure" search problem.

The idea of using \acp{GA} to find a polynomial regression is not new \cite{maertens2006genetic, yu2008optimal, wu2009novel} but still generates original research \cite{hofwing2011optimal, cetisli2011polynomial}. In line with that research this work describes a general method to find a polynomial regression of a given dataset. The optimal regression minimizes a cost function that accounts for both the \ac{rmse} and a regularization factor to avoid over-fitting.

@@

\item find the fittest set of monomials, \emph{i.e.} the polynomial that minimizes the \ac{rmse} on the same dataset;
\end{enumerate}
%
More precisely, concerning the first problem, let $\mathcal{D}$ be a dataset with $n$ observations of variables $Y, X_1,\ldots,X_m$ and $\mathcal{P} = \left\lbrace q_1,\ldots, q_k\right\rbrace$ a set of $k$ monomial expressions over $X_1,\ldots,X_m$. Define the hypothesis\footnote{The expression $q|_{X=x}$ reads ``\emph{replace all instances of $X$ by $x$ in $q$}''.}
%
\begin{eqnarray*}
h_{\Theta,\mathcal{P}}\left(x_1,\ldots,x_m\right) &=& \sum_{j = 1}^k \theta_j q_j|_{X_i=x_i,\forall 1 \leq i \leq m}
\end{eqnarray*}
%
and let the cost
%
\begin{eqnarray}
J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right) &=& \sqrt{\frac{1}{n}\sum_{i=1}^n \left( y\sample{i} - h_{\Theta,\mathcal{P}}\left( x_1\sample{i},\ldots,x_m\sample{i} \right) \right)^2 }\label{eq:rmse}
\end{eqnarray}
%
be the usual \acf{rmse} function. Now the first problem can be stated as:
%
\begin{quote}
Given a dataset $\mathcal{D}$ and a set of monomials $\mathcal{P}$, find parameters $\Theta$ that minimize $J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right)$.
\end{quote}
%
It turns out that this problem can be solved as a usual linear regression problem by expanding $\mathcal{D}$ with columns that replicate the monomials in $\mathcal{M}$.

The second problem is treated in the \ac{GA} setting: Let $\mathcal{D}$ be a dataset as above and $Q$ a set of polynomials. For each polynomial $p\in Q$ let $\mathcal{P}_p$ be the set of monomials in $p$ (without the coefficients) and compute the fitness
%
\begin{eqnarray*}
\phi_p &=& \min_\Theta J_{\textrm{fit}}\left(\Theta;\mathcal{P}_p,\mathcal{D}\right)
\end{eqnarray*}
%
by solving the first problem. With a fitness of every instance, a \ac{GA} will apply genetic operators (usually mutation and crossover) to evolve the population $Q$ until a reasonable approximation of a local minimum is found. 
%
\begin{algorithm}[tb]
\begin{algorithmic}
\Function{GAPoly}{$D,pop_0,\epsilon$}
	\State $pop \gets pop_0;\: err \gets 1.0+\epsilon$
	\While{$err > \epsilon$}
		\State $pop \gets \Call{IterateGA}{pop}$\Comment{Build next generation}
		\State $pop \gets \Call{Sort}{pop, key = \textsc{Lm.rmse}}$\Comment{Sort by \ac{rmse} of linear regression}
		\State $err \gets \textsc{rmse}\left(\Call{First}{pop}\right)$
	\EndWhile
	\State\Return $\Call{First}{pop}$
\EndFunction
\end{algorithmic}
\caption{\ac{GAPoly} uses linear regression to find monomial coefficients that minimize the \ac{rmse} over a dataset and \acp{GA} to explore the space of polynomials. At exit the \ac{rmse} of the fittest instance is bounded by $\epsilon$.}\label{alg:gapoly}
\end{algorithm} 
%
Notice that the properties of \acp{GA} and linear regression entail that
the composition of \acp{GA} with linear regression, as defined in Algorithm \ref{alg:gapoly}, converges to a polynomial that is a local minimum of the fitness function, encapsulated in the \ac{rmse} function $J_{\textrm{fit}}$.
%

@@

This dataset was used in order to verify if \ac{GAPoly} was able to find the polynomial relation, which the algorithm did (cf. figure~\ref{artificial_dataset1_lambda1.0}). The genetic algorithm run with a population of $n=100$~solutions with $50$~iterations for each run.

@@

\item{\textsc{Housing}}: This data set concerns the task of predicting housing values in areas of Boston. There are $m=13$ continuous attributes and the dependent variable is the median value of owner-occupied homes in \$1000's. There are $n=506$ observations.

@@

The proposed method has competitive results comparing with some well-known regression methods. Only Random Forest outperforms \ac{GAPoly} systematically (it also outperforms all the other regression algorithms). One exception --- the artificial dataset that uses a straightforward polynomial relation --- is the Auto-MPG dataset where \ac{GAPoly} has comparable results. This is evidence that applying standard genetic operators for polynomial model searching is a viable tool for regression purposes.

For complexity considerations \ac{GAPoly} demands some processing time. On a current quad-core computer, processing the Kinematics dataset (with 8k~observations) takes approximately 5~minutes. The processing time can probably be speeded by one to two orders in magnitude if the process is implemented in a low level programming language like C++. However, speed optimization was not the focus of this article.

@@

The authors are grateful to the \textsc{Fundação para a Ciência e Tecnologia} (FCT) and the  R\&D laboratory \textsc{LabMAg} for the financial support given to this work, under the strategic project \textsc{PEst-OE/EEI/UI0434/2011}.

The datasets used herein were selected from Luis Torgo's data  repository, \url{http://www.dcc.fc.up.pt/~ltorgo/Regression/DataSets.html}. Most come originally from UCI ML repository, \url{http://archive.ics.uci.edu/ml/}.
